---
title: "Intro to Transcriptomics"
author: Nicole Black, Wade Boohar, Kayla Xu
date: 07/17/22
updated: 10/18/24
---

***Deliverables***
-Upload this R Notebook to your GitHub and submit the link to your Repo on Brightspace.
-Include any graphs or figures created in this assignment in the folder with your R notebook with descriptive file names.

Since this is an optional partner activity, it is okay if your answers are the same as your partner’s as long as everyone understands it and could explain it in their own words if asked. Each person must individually push their code to Github. *At the top of your R Notebook, write the name of you and your partner(s) as a comment.*

***Complete the following coding activity and answer any following questions as comments in your R Notebook***

In SummarizedExperiment Tutorial, you learned how to manipulate the SummarizedExperiment data structure and turn it into more readable dataframes, saving them as rna_counts, rna_clinical, and rna_genes. In this semi-guided assignment, you will use these dataframes to perform differential expression analysis based on tumor status.

*Pre-Assignment*
Use knitr function to set your working directory to your analysis_data folder in 490_cluster.
```{r setup}
 knitr::opts_knit$set(root.dir = normalizePath("/home1/aditis54/490_cluster/analysis_data"))
```

If DESeq2 is not already installed, install it now
```{r}
if (!require("DESeq2", quietly = TRUE))
BiocManager::install("DESeq2")
BiocManager::install("EnhancedVolcano") 
library(EnhancedVolcano)
```

Load in all necessary packages
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)

```



*1*
Read in the rna_clinical, rna_genes, and rna_counts dataframes which you made in the "SummarizedExperiment Guided Tutorial" R Notebook

```{r}
rna_query <- GDCquery(project ="TCGA-BRCA",
                     data.category = "Transcriptome Profiling",
                     data.type = "Gene Expression Quantification",
                      workflow.type = "STAR - Counts")

rna_se <- GDCprepare(rna_query)

rna_clinical <- rna_se@colData
rna_clinical <- as.data.frame(rna_clinical)
colnames(rna_clinical)
rna_clinical <- rna_clinical[, -25] # removes treatments

rna_genes <- rna_se@rowRanges@elementMetadata
rna_genes <- as.data.frame(rna_genes)
row.names(rna_genes) <- rna_genes$gene_id # add descriptive row name

```
```{r}
rna_counts <- rna_se@assays@data$unstranded
rna_counts <- as.data.frame(rna_counts)
row.names(rna_counts) <- rna_genes$gene_id # add descriptive row name
colnames(rna_counts) <- rna_clinical$barcode
```


*2*
In this assignment, you will run differential expression analysis comparing patient samples by whether the sample is from a tumor or normal tissue (this is the definition column in rna_clinical). You will need to choose a variable to control for covariance of: age and/or PAM50 subtype (paper_BRCA_Subtype_PAM50). 

Manipulate those columns so that they are ready for differential expression analysis (hint: what kind of variables are they? what data type are they by default? do you need to handle unknown values?) Filter out genes with a total expression across all patients less than 1000.

```{r}
# variable of interest
definition_mask <- ifelse(rna_clinical$definition == "Metastatic", F, T)
rna_clinical <- rna_clinical[definition_mask, ]
rna_clinical$definition <- factor(rna_clinical$definition)
```

```{r}
# covariates


NA_age_mask <- !is.na(rna_clinical$age_at_diagnosis)
rna_clinical <- rna_clinical[NA_age_mask, ]
# rna_clinical$age_at_diagnosis <- factor(rna_clinical$age_at_diagnosis)

design <- ~age_at_diagnosis + definition


# NA_PAM50_mask <- !is.na(clean_clinical$paper_BRCA_Subtype_PAM50)
# clean_clinical <- clean_clinical[NA_PAM50_mask, ]
# clean_clinical$paper_BRCA_Subtype_PAM50 <- factor(clean_clinical$paper_BRCA_Subtype_PAM50)

# prefiltering genes
# keep_genes <- rowSums(rna_counts) >= 1000
# filtered_rna_counts <- rna_counts[keep_genes, ]

```

```{r}
# matching the clinical, counts, and genes dataframes
clean_counts <- rna_counts[, rna_clinical$barcode, drop = FALSE]
# clean_counts variables should match rna_clinical obs

less_1000_mask <- rowSums(rna_counts) >= 1000
clean_counts <- clean_counts[less_1000_mask, ]
clean_genes <- rna_genes[less_1000_mask, ]
```

```

*3*
Perform the differential expression analysis, All you need to do is fill in the appropriate # terms
```{r}
dds <- DESeqDataSetFromMatrix(countData = clean_counts,
                              colData = rna_clinical,
                              design=~age_at_diagnosis + definition)

dds_obj <- DESeq(dds)

resultsNames(dds_obj)

results <- results(dds_obj, format = "DataFrame", contrast = c("definition", "Primary solid Tumor" , "Solid Tissue Normal"))

results <- data.frame(results)
```

Prepare results dataframe for EnhancedVolcano plotting. Add two columns, "-log10(padj)" and "gene_name". Fill in these columns appropriately.
```{r}
results$gene_name <- clean_genes$gene_name
results$"-log10(padj)" <- -log10(results$padj) 
```

*4*
Now we will use the EnhancedVolcano package to plot our results. The code is already completed and should run without adjustment if all code up to here is correct.
```{r}
EnhancedVolcano(results,
                lab = results$gene_name,
                x = 'log2FoldChange',
                y = '-log10(padj)',
                title = 'Sample Definition: Tumor vs Normal Tissue',
                pointSize = 1.0,
                labSize = 5.0)
```

*5*
# Explain what genes from each part of the Volcano Plot mean in terms of their significance and up/down regulation. 

top-right genes: The top-right genes in the plot would generally mean that the specific gene is significantly up regulated in Tumor patients compared to normal tissue patients. In our plot however, there do not seem to be any top-right genes.

bottom-right genes: The bottom-right genes in the plot would generally mean that the specific gene is not significantly up regulated in Tumor patients compared to normal tissue patients.

top-left genes: The top left genes are significantly down regulated in Tumor patients compared to normal tissue patients.

bottom-left genes: The bottom left genes is not significantly down regulated in Tumor patients compared to normal tissue patients.

top-middle genes: high P-values, fold changes are close to zero so there is no substantial upregulation or downregulation 

bottom-middle genes: fold changes are close to zero and they have low P-values, and they are still neither significantly upregulated nor downregulated.

Save the picture of the volcano plot (using either ggsave() or right clicking and manually downloading the image and push this .Rmd and the image to GitHub)